---
title: "Atlas analysis affter SoupX and filtering"
date: "`r format(Sys.time(),  '%d %B %Y')`"
author: "Mohamed Hassan"
output: 
  html_notebook:
    theme: cerulean
    toc: true
    toc_depth: 3
editor_options: 
  markdown: 
    wrap: 72
---



```{r}
library(tidyverse)
library(Seurat)
```


```{r}
In_House_SAT <- readRDS("In_House_after_SoupX_filt_and_doublets.RDS")
In_House_SAT
```


################################################################################
################################################################################

# Analysis


```{r}
DefaultAssay(In_House_SAT) <- "RNA"
```

## Adding Cell cycle score
```{r}
library("RCurl")
library(AnnotationHub)

cc_file <- getURL("https://raw.githubusercontent.com/hbc/tinyatlas/master/cell_cycle/Homo_sapiens.csv") 
cell_cycle_genes <- read.csv(text = cc_file)


# Connect to AnnotationHub
ah <- AnnotationHub()

# Access the Ensembl database for organism
ahDb <- query(ah, 
              pattern = c("Homo sapiens", "EnsDb"), 
              ignore.case = TRUE)

# Acquire the latest annotation files
id <- ahDb %>%
        mcols() %>%
        rownames() %>%
        tail(n = 1)

# Download the appropriate Ensembldb database
edb <- ah[[id]]

# Extract gene-level information from database
annotations <- genes(edb, 
                     return.type = "data.frame")

# Select annotations of interest
annotations <- annotations %>%
        dplyr::select(gene_id, gene_name, seq_name, gene_biotype, description)



# Get gene names for Ensembl IDs for each gene
cell_cycle_markers <- dplyr::left_join(cell_cycle_genes, annotations, by = c("geneID" = "gene_id"))

# Acquire the S phase genes
s_genes <- cell_cycle_markers %>%
        dplyr::filter(phase == "S") %>%
        pull("gene_name")
        
# Acquire the G2M phase genes        
g2m_genes <- cell_cycle_markers %>%
        dplyr::filter(phase == "G2/M") %>%
        pull("gene_name")

In_House_SAT <- CellCycleScoring(In_House_SAT, 
                                   s.features = s_genes, 
                                   g2m.features = g2m_genes)
```

## Quality check
```{r}
FeatureScatter(In_House_SAT, 
               feature1 = "nCount_RNA", 
               feature2 = "nFeature_RNA", 
               group.by = "orig.ident", raster = F)
```


```{r}
In_House_SAT <- subset(In_House_SAT, 
                   nFeature_RNA < 10000 &
                   nCount_RNA < 60000)
In_House_SAT
```

```{r}
FeatureScatter(In_House_SAT, 
               feature1 = "nCount_RNA", 
               feature2 = "nFeature_RNA", 
               group.by = "orig.ident", raster = F)
```

## Linear model
```{r}
lin_mod <- lm(log(In_House_SAT@meta.data$nFeature_RNA) ~ log(In_House_SAT@meta.data$nCount_RNA))
sub <- lin_mod$model[lin_mod$residuals < -0.5,]
plot (log(In_House_SAT@meta.data$nCount_RNA),
      log(In_House_SAT@meta.data$nFeature_RNA), 
      xlab = "log (nCount_RNA)",
      ylab = "log (nFeature_RNA)",
      pch=20,col="grey")
abline (lin_mod,col="red",lwd=3)
points (sub[,2], sub[,1],pch=20,col="blue")
tokeep <- as.numeric(names(lin_mod$residuals[lin_mod$residuals >= -0.5]))
In_House_SAT <- subset(In_House_SAT, cells = tokeep)
```

```{r}
FeatureScatter(In_House_SAT, 
               feature1 = "nCount_RNA", 
               feature2 = "nFeature_RNA", 
               group.by = "orig.ident", raster = F)
```


# 2. Data Processing
```{r}
DefaultAssay(In_House_SAT)<- "RNA"
In_House_SAT <- NormalizeData(In_House_SAT, normalization.method = "LogNormalize", 
                                scale.factor = 10000)
In_House_SAT <-  FindVariableFeatures(In_House_SAT, selection.method = "vst", 
                                        nfeatures = 2000)
In_House_SAT <-  ScaleData(In_House_SAT, vars.to.regress =  "nCount_RNA")
```


```{r}
In_House_SAT <-  RunPCA(In_House_SAT, features = VariableFeatures(object = In_House_SAT), ndims.print = 1:40)
```


```{r}
ElbowPlot(In_House_SAT, ndims = 50)

### Determining the number of significant PCs
pct <- In_House_SAT@reductions$pca@stdev / sum(In_House_SAT@reductions$pca@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
sig_pcs <- 1:pcs
```


```{r}
In_House_SAT <- FindNeighbors(In_House_SAT,
                          reduction = "pca", dims = sig_pcs)
In_House_SAT <- RunUMAP(In_House_SAT, dims = sig_pcs)
```

```{r, fig.width=21, fig.height=7}
DimPlot(In_House_SAT, label = T, 
         group.by = c("orig.ident", "Phase"), 
         raster = F) + NoLegend()
```


```{r, fig.width=16,  fig.height=8, eval=FALSE}
UMAPPlot(In_House_SAT, label = T, 
         group.by = c("orig.ident","Phase"), 
         raster = F)

pdf("/home/lucamannino/R_Projects/scAtlas/SAT/UMAP_before_batch_correction.pdf", width = 12, height = 4)
UMAPPlot(In_House_SAT, label = T, 
         group.by = c("orig.ident", "Condition", "Phase"), 
         raster = F) + NoLegend()
dev.off()
```



```{r, eval=FALSE}
saveRDS(In_House_SAT, "R_Projects/scAtlas/SAT/In_House_SAT_UnIntegratedObj.RDS")
```


# 3. Integration (Removing batch effect)
```{r}
library(harmony)
In_House_SAT$orig.ident <- as.factor(In_House_SAT$orig.ident)

## Adding samples column for integration
In_House_SAT$samples<- as.character(In_House_SAT$orig.ident)

# Running the integration with Harmony
SAT_IntegratedObj <- RunHarmony(In_House_SAT, 
                             group.by.vars= "samples", 
                             verbose = T, assay.use= "RNA",
                             max.iter.harmony = 100)

ElbowPlot(SAT_IntegratedObj, reduction =  "pca", ndims = 50) +  ElbowPlot(SAT_IntegratedObj, reduction =  "harmony", ndims = 50)


### Determining the number of significant PCs
pct <- SAT_IntegratedObj@reductions$harmony@stdev / sum(SAT_IntegratedObj@reductions$harmony@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
harmony_sig_pcs <- 1:pcs
harmony_sig_pcs
```


```{r}
SAT_IntegratedObj <- RunUMAP(SAT_IntegratedObj, 
                             dims = harmony_sig_pcs, 
                             reduction = "harmony", verbose = T)

# SAT_IntegratedObj <- RunUMAP(SAT_IntegratedObj, 
#                              dims = c(1:30), 
#                              reduction = "harmony", verbose = T)
```


## Checking the quality of integration
```{r}
UMAPPlot(SAT_IntegratedObj, group.by = "orig.ident", raster= F) + NoLegend()
```


```{r, eval=FALSE}
pdf("SAT_IntegratedObj_sample_Integration_quality.pdf")
UMAPPlot(SAT_IntegratedObj, group.by = "orig.ident", raster= F) + NoLegend()
dev.off()
```



```{r, fig.height = 10}
# UMAP split by study
DimPlot(SAT_IntegratedObj, split.by ="orig.ident",group.by= "orig.ident", raster = F,
       label = T, repel = F, ncol = 5) & NoLegend()
```



```{r}
SAT_IntegratedObj<- FindNeighbors(SAT_IntegratedObj, dims = harmony_sig_pcs, 
                                    reduction = "harmony", verbose = FALSE)
res <- seq(0.1, 0.4, by=0.1)
SAT_IntegratedObj <- FindClusters(SAT_IntegratedObj, verbose = T, 
                                    resolution = res)
```


## Selecting optimal resolution based on clustree
```{r, eval=FALSE, fig.width=10, fig.height=10}
library(clustree)
Clustree_Plot <- clustree::clustree(SAT_IntegratedObj)
Clustree_Plot
```


```{r, fig.width=20, fig.height=5}
UMAPPlot(SAT_IntegratedObj, 
         group.by = grep(pattern = "^RNA_snn", 
                         x = colnames(SAT_IntegratedObj@meta.data)), 
         label = T, ncol = 4, raster = F) & NoLegend()
```

```{r, fig.width=10}
DotPlot(SAT_IntegratedObj, 
        features = c("PECAM1", "CDH5", "VWF", "MECOM",
                     "MMRN1", "PROX1", "MYH11", 
                     "CARMN", "STEAP4", "MRC1", "MSR1"),
        group.by = "RNA_snn_res.0.2")
```

```{r}
DimPlot(SAT_IntegratedObj, group.by = "RNA_snn_res.0.2", raster = F, label = T, repel = T)
```

# 2. Subsetting the Vascular cells (Endothelial cells, pericytes, and VSMCs)
```{r}
Idents(SAT_IntegratedObj) <- SAT_IntegratedObj$RNA_snn_res.0.2
VascularObj <- subset(SAT_IntegratedObj, 
                      RNA_snn_res.0.2 %in% c("3", "4", "5", "6", "9"))
VascularObj
```


```{r}
UMAPPlot(VascularObj, label = T)
```

```{r, fig.width=10, fig.height=12}
FeaturePlot(VascularObj, features = c("CDH5", "PECAM1", "MYH11", "RGS6", "PROX1", "MMRN1"), order = T)
```



```{r}
DefaultAssay(VascularObj) <- "RNA"
```


## Preprocessing the Vascular object
```{r}
VascularObj <- NormalizeData(VascularObj, normalization.method = "LogNormalize", 
                             scale.factor = 10000)
VascularObj <-  FindVariableFeatures(VascularObj, selection.method = "vst", nfeatures = 2000)
VascularObj <-  ScaleData(VascularObj, vars.to.regress="nCount_RNA")
VascularObj<-  RunPCA(VascularObj, ndims.print = 1:50)
```


```{r}
ElbowPlot(VascularObj, ndims = 50)
```


```{r}
### Determining the number of significant PCs
pct <- VascularObj@reductions$pca@stdev / sum(VascularObj@reductions$pca@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
sig_pcs <- 1:pcs
```


```{r}
VascularObj <- FindNeighbors(VascularObj, dims = sig_pcs)
```

```{r}
VascularObj <- RunUMAP(VascularObj, dims = sig_pcs)
UMAPPlot(VascularObj, group.by = "orig.ident")
```


```{r}
library(harmony)
VascularObj$samples <- as.factor(VascularObj$orig.ident)
VascularObj$samples <- as.character(VascularObj$samples)

Vascular_IntegratedObj <- RunHarmony(object = VascularObj,
                                     group.by.vars= "samples",
                                     verbose = T, assay.use= "RNA",max.iter.harmony = 100)

### Determining the number of significant PCs
pct <- Vascular_IntegratedObj@reductions$harmony@stdev / sum(Vascular_IntegratedObj@reductions$harmony@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
harmony_sig_pcs <- 1:pcs
harmony_sig_pcs

Vascular_IntegratedObj <- RunUMAP(Vascular_IntegratedObj,
                                  dims = harmony_sig_pcs,
                                  reduction = "harmony", verbose = T)
```

## Quality of the integration
```{r}
UMAPPlot(Vascular_IntegratedObj, group.by = "orig.ident") + NoLegend()
```


```{r}
Vascular_IntegratedObj <- FindNeighbors(Vascular_IntegratedObj,
                                        dims = harmony_sig_pcs,
                                        reduction = "harmony", verbose = T)
```


```{r, fig.width=12}
res <- seq(0.1, 1, 0.1)
Vascular_IntegratedObj <- FindClusters(Vascular_IntegratedObj,resolution = res)
# library(clustree)
# clustree(Vascular_IntegratedObj)
```


```{r, fig.width=20, fig.height=10}
UMAPPlot(Vascular_IntegratedObj, 
         group.by = grep(pattern = "^RNA_snn", 
                         x = colnames(Vascular_IntegratedObj@meta.data)), 
         label = T, ncol = 5) & NoLegend()
```


## Signature plots
```{r}
ECs_markers_ref <- xlsx::read.xlsx("/home/lucamannino/Downloads/Vascular_markers_summary.xlsx", sheetIndex = 1)
ECs_markers_ref <- separate(ECs_markers_ref, col = "Capillary.ECs", 
                            into = c("Capillary.ECs", NA), sep = " ")

ECs_markers_ref <- lapply(ECs_markers_ref, function(x){
  x <- x[!(is.na(x))]
})
```


## Assiging signatures
```{r}
for (cell_type in names(ECs_markers_ref)) {
  
  # Access gene list for the current cell type
  current_gene_list <- ECs_markers_ref[[cell_type]]
  
  # Run AddModuleScore function on the Seurat object for the current cell type's gene list
  Vascular_IntegratedObj <- AddModuleScore(object = Vascular_IntegratedObj,
                               features = list(current_gene_list),
                               name = str_c(cell_type, "signature", sep = "_"),
                               search = T)
}
```


```{r}
colnames(Vascular_IntegratedObj@meta.data) <- gsub(pattern = "_signature1$", 
                                                   replacement = "_signature", 
                                                   x=colnames(Vascular_IntegratedObj@meta.data))
```

```{r}
pal <- viridis::viridis(n = 10, option = "C", direction = -1)
```


## Plotting signatures
```{r, fig.width=14, fig.height=14}
signature_columns <- colnames(Vascular_IntegratedObj@meta.data[grep(pattern = "_signature$",
                                                                    x = colnames(Vascular_IntegratedObj@meta.data))])

FeaturePlot(Vascular_IntegratedObj, 
         features = signature_columns,
         cols = pal, ncol = 3)
```
```{r}
pdf("/home/lucamannino/R_Projects/scAtlas/Revision/Ambient_RNA/SoupX_plots.pdf", width = 10, height = 10)
DimPlot(SAT_IntegratedObj, raster = F, group.by = "RNA_snn_res.0.2") + ggtitle("") + NoAxes()
DimPlot(SAT_IntegratedObj, raster = F, group.by = "RNA_snn_res.0.2") + ggtitle("") + NoAxes() + NoLegend()
DimPlot(Vascular_IntegratedObj, group.by = "RNA_snn_res.0.2") + ggtitle("") + NoAxes() + NoLegend()
FeaturePlot(Vascular_IntegratedObj, 
            features = signature_columns[length(signature_columns)],
            cols = c("grey", "red4"))
FeaturePlot(Vascular_IntegratedObj, 
            features = signature_columns[length(signature_columns)],
            cols = c("grey", "red4")) + NoAxes() + NoLegend() + ggtitle("")
dev.off()
```


```{r, fig.height = 8}
FeaturePlot(Vascular_IntegratedObj, features = c("PECAM1", "CDH5", "VWF", "MECOM"))
```

```{r}
DotPlot(Vascular_IntegratedObj, features = c("MRC1", "CFD", "IL7R", "PLIN1"))
```

```{r, fig.height = 8}
FeaturePlot(Vascular_IntegratedObj, features = c("MRC1", "CFD", "IL7R", "PLIN1"))
```



```{r, eval=FALSE}
DefaultAssay(Vascular_IntegratedObj) <- "RNA"
Idents(Vascular_IntegratedObj) <- Vascular_IntegratedObj$RNA_snn_res.0.3

markers_vascular_res03 <- FindAllMarkers(Vascular_IntegratedObj,
                              only.pos = T,
                              min.pct = 0.15,
                              logfc.threshold = 0.3)

# writexl::write_xlsx(as.data.frame(SAT_markers_res09), "markers_Atlas_WRST_15PCT_Log03_res09.xlsx")
```



