---
title: "scds"
date: "`r format(Sys.time(),  '%d %B %Y')`"
author: "Mohamed Hassan"
output: 
  html_notebook:
    theme: cerulean
    toc: true
    toc_depth: 3
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = T, results = "hide")
require("knitr")
opts_knit$set(root.dir = "/home/mhassan/")
```

```{r random-seed-setting}
set.seed(12345)
```

```{r loading-libraries, results='hide'}
pacman::p_load(tidyverse, Seurat, patchwork, scds, SingleCellExperiment)
```

```{r}
In_House <- readRDS("/home/mhassan/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/InHouse_merged.RDS")
In_House
```


```{r}
sce <- as.SingleCellExperiment(In_House)
```


# Run scds

```{r}
## RUn co-expression baed graph doublet scoring
sce <- cxds(sce, retRes = TRUE, estNdbl=TRUE)
```

```{r}
#- Annotate doublet using binary classification based doublet scoring:
sce <- bcds(sce, retRes = TRUE, estNdbl=TRUE)
```

```{r}
#- Combine both annotations into a hybrid annotation
sce <- cxds_bcds_hybrid(sce, estNdbl=TRUE)
```


```{r}
#- Doublet scores are now available via colData:
CD  = colData(sce)
head(cbind(CD$cxds_score,CD$bcds_score, CD$hybrid_score))

```


```{r}
summary <- as.data.frame(table(Doublets$scds_DropletType))
colnames(summary) <- c("Classification", "Droplet N")
write_delim(summary, 
            "/home/mhassan/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/scds_doublets_singlets.tsv")
```


```{r}
Doublets <- as.data.frame(cbind(rownames(colData(sce)), 
                                colData(sce)$hybrid_score, 
                                colData(sce)$hybrid_call))


## Doublet scores are now available via colData:
colnames(Doublets) <- c("Barcode","scds_score","scds_DropletType")
Doublets$scds_DropletType <- gsub("FALSE","singlet",Doublets$scds_DropletType)
Doublets$scds_DropletType <- gsub("TRUE","doublet",Doublets$scds_DropletType)

write_tsv(Doublets, 
          "/home/mhassan/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/scds_doublets_singlets.tsv")
```

```{r}
head(Doublets)
```

```{r}
In_House <- AddMetaData(In_House, metadata = Doublets)
```

```{r}
In_House@meta.data %>% 
  as.data.frame() %>% 
  write_tsv("/home/mhassan/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/InHouse_df_scds_res.tsv")
```


```{r}
In_House$cxds_score <- sce$cxds_score
In_House$cxds_call <- sce$cxds_call
In_House$bcds_score <- sce$bcds_score
In_House$bcds_call <- sce$bcds_call
In_House$hybrid_score <- sce$hybrid_score
In_House$hybrid_call <- sce$hybrid_call
```


## Plots for the paper 
```{r}
hist(as.numeric(In_House$scds_score))
```

```{r}
hist(as.numeric(In_House$cxds_score))
```


```{r}
hist(as.numeric(In_House$bcds_score))
```


```{r}
hist(as.numeric(In_House$hybrid_score))
```


```{r}
obj_before_doub_rem <- In_House
obj_before_doub_rem
```

## Filtering doublets
```{r}
In_House
In_House <- subset(In_House, scds_DropletType == "singlet")
In_House
```

Number of doublets removed are `r ncol(obj_before_doub_rem) - ncol(In_House)`

```{r}
FeatureScatter(In_House,
               feature1 = "nCount_RNA",
               feature2 = "nFeature_RNA", 
               group.by = "orig.ident", raster = F)
```


## Filtering outliers
```{r}
In_House
In_House <- subset(In_House, 
                   nCount_RNA < 20000 &
                   nFeature_RNA < 10000)
In_House
```


```{r}
FeatureScatter(In_House,
               feature1 = "nCount_RNA",
               feature2 = "nFeature_RNA", 
               group.by = "orig.ident", raster = F)
```


# Adding metadata information
```{r}
library(dplyr)
library(tibble)

# Sample-level metadata
sample_metadata <- tribble(
  ~orig.ident, ~Condition,         ~gender,  ~BMI,  ~Age,
  "SAT1",      "Obese",            "male",   41.7,  43,
  "SAT4",      "Obese",            "male",   49.7,  30,
  "SAT9",      "Obese",            "female", 39.5,  37,
  "SAT10",     "Diabetic Obese",   "female", 61.5,  41,
  "SAT13",     "Lean",             "female", 23.0,  42,
  "SAT17",     "Diabetic Obese",   "female", 52.6,  34,
  "SAT18",     "Obese",            "female", 61.3,  29,
  "SAT22",     "Diabetic Obese",   "male",   39.7,  50,
  "SAT34",     "Obese",            "male",   62.5,  40,
  "SAT39",     "Obese",            "male",   61.2,  41,
  "SAT51",     "Obese",            "female", 57.0,  44,
  "SAT54",     "Lean",             "female", 22.3,  61,
  "SAT58",     "Lean",             "male",   24.3,  54,
  "SAT60",     "Lean",             "male",   23.5,  26
)

# Convert to lookup table indexed by sample ID
sample_metadata_df <- sample_metadata %>%
  column_to_rownames("orig.ident")

# Match sample metadata to each cell using orig.ident
cell_metadata <- sample_metadata_df[In_House$orig.ident, , drop = FALSE]
rownames(cell_metadata) <- colnames(In_House)
```


```{r}
# Add metadata back to Seurat object
In_House <- AddMetaData(In_House, metadata = cell_metadata)
```


################################################################################
################################################################################


# Analysis


```{r}
DefaultAssay(In_House) <- "RNA"
```

## Adding Cell cycle score
```{r}
In_House <- CellCycleScoring(In_House, 
                                   s.features = Seurat::cc.genes$s.genes, 
                                   g2m.features = Seurat::cc.genes$g2m.genes)
```


## Data Processing
```{r}
DefaultAssay(In_House)<- "RNA"
In_House <- NormalizeData(In_House, normalization.method = "LogNormalize", 
                                scale.factor = 10000)
In_House <-  FindVariableFeatures(In_House, selection.method = "vst", 
                                        nfeatures = 2000)
In_House <-  ScaleData(In_House)
```


```{r}
In_House <-  RunPCA(In_House, features = VariableFeatures(object = In_House), ndims.print = 1:40)
```


```{r}
ElbowPlot(In_House, ndims = 50)

### Determining the number of significant PCs
pct <- In_House@reductions$pca@stdev / sum(In_House@reductions$pca@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
sig_pcs <- 1:pcs
sig_pcs
```


```{r}
In_House <- FindNeighbors(In_House,
                          reduction = "pca", dims = c(1:30))
In_House <- RunUMAP(In_House, dims = c(1:30))
```


```{r, fig.width=21, fig.height=7}
DimPlot(In_House, 
         group.by = c("orig.ident", "Phase"), 
         raster = F) + NoLegend()
```

```{r}
DimPlot(In_House, 
         group.by = "orig.ident", 
         raster = F)
```



```{r, eval=FALSE}
saveRDS(In_House, "~/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/In_House_CB_scds_UnInt.RDS")
```


# 3. Integration (Removing batch effect)
```{r}
library(harmony)
In_House$orig.ident <- as.factor(In_House$orig.ident)

## Adding samples column for integration
In_House$samples<- as.character(In_House$orig.ident)

# Running the integration with Harmony
In_House_Integrated <- RunHarmony(In_House, 
                             group.by.vars= "samples", 
                             verbose = T, assay.use= "RNA",
                             max.iter.harmony = 100)

ElbowPlot(In_House_Integrated, reduction =  "pca", ndims = 50) +  ElbowPlot(In_House_Integrated, reduction =  "harmony", ndims = 50)


### Determining the number of significant PCs
pct <- In_House_Integrated@reductions$harmony@stdev / sum(In_House_Integrated@reductions$harmony@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
harmony_sig_pcs <- 1:pcs
harmony_sig_pcs
```


```{r}
In_House_Integrated <- RunUMAP(In_House_Integrated, 
                             dims = c(1:30),
                             reduction = "harmony", verbose = T)
```


## Checking the quality of integration
```{r}
DimPlot(In_House_Integrated, group.by = "orig.ident", raster= F)
```



```{r, fig.height = 10}
# UMAP split by sample
DimPlot(In_House_Integrated, split.by ="orig.ident",group.by= "orig.ident", raster = F,ncol = 5)
```



```{r}
In_House_Integrated<- FindNeighbors(In_House_Integrated, dims = c(1:30), 
                                    reduction = "harmony", verbose = FALSE)
res <- seq(0.1, 1, by=0.1)
In_House_Integrated <- FindClusters(In_House_Integrated, verbose = T, 
                                    resolution = res)
```


## Selecting optimal resolution based on clustree
```{r, eval=FALSE, fig.width=10, fig.height=10}
library(clustree)
Clustree_Plot <- clustree::clustree(In_House_Integrated)
Clustree_Plot
```




```{r}
In_House_Integrated@meta.data <- In_House_Integrated@meta.data %>% 
  relocate(RNA_snn_res.0.8, .before = RNA_snn_res.0.9)
```


```{r, fig.width=20, fig.height=15}
DimPlot(In_House_Integrated, 
         group.by = grep(pattern = "^RNA_snn", 
                         x = colnames(In_House_Integrated@meta.data)), 
         label = T, ncol = 4, raster = F) & NoLegend()
```


```{r, fig.width=10}
DotPlot(In_House_Integrated, 
        features = c("PECAM1", "CDH5", "VWF", "MECOM",
                     "MMRN1", "PROX1", "MYH11", 
                     "CARMN", "STEAP4", "MRC1", "IL7R", "APOD", "PLIN1", 
                     "MSR1"),
        group.by = "RNA_snn_res.0.1")
```


```{r}
DimPlot(In_House_Integrated, 
        group.by = "RNA_snn_res.0.1", 
        raster = F, label = T, 
        repel = T)
```

```{r}
Idents(In_House_Integrated) <- In_House_Integrated$RNA_snn_res.0.1
top_markers_res01 <- FindAllMarkers(In_House_Integrated, only.pos = T)
```



```{r}
saveRDS(In_House_Integrated, "/home/mhassan/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/In_House_Integrated.RDS")
```



```{r}
new_names = c(
'0' = 'ASPCs',
'1' = 'Macro_Mono',
'2' = 'ECs',
'3' = 'Adipocytes',
'4' = 'Mural',
'5' = 'T_NK',
'6' = 'ECs',
'7' = 'Mast',
'8' = 'LECs',
'9' = 'Prolif'
)

Idents(In_House_Integrated) <- In_House_Integrated$RNA_snn_res.0.1

In_House_Integrated$celltype_annotation <- recode(Idents(In_House_Integrated),
                                                  !!!new_names)
```


```{r}
DimPlot(In_House_Integrated, group.by = "celltype_annotation", label = T, repel = T, raster = F) + NoLegend()
```


```{r}
Idents(In_House_Integrated) <- "celltype_annotation"
markers_main_aft_ann <- FindAllMarkers(In_House_Integrated, only.pos = T)
```


```{r}
writexl::write_xlsx(markers_main_aft_ann, "~/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/MainObj_markers_aft_ann.xlsx")
```





# 2. Subsetting the Vascular cells (Endothelial cells, pericytes, and VSMCs)
```{r}
Idents(In_House_Integrated) <- In_House_Integrated$celltype_annotation
VascularObj <- subset(In_House_Integrated, 
                      celltype_annotation %in% c("Mural", "ECs", "LECs"))
VascularObj
```


```{r}
UMAPPlot(VascularObj, label = T)
```

```{r, fig.width=10, fig.height=12}
FeaturePlot(VascularObj, features = c("CDH5", "PECAM1", "MYH11", "RGS6", "PROX1", "MMRN1"), order = T)
```



```{r}
DefaultAssay(VascularObj) <- "RNA"
```


## Preprocessing the Vascular object
```{r}
VascularObj <- NormalizeData(VascularObj, normalization.method = "LogNormalize", 
                             scale.factor = 10000)
VascularObj <-  FindVariableFeatures(VascularObj, selection.method = "vst", nfeatures = 2000)
VascularObj <-  ScaleData(VascularObj)
VascularObj<-  RunPCA(VascularObj, ndims.print = 1:50)
```


```{r}
ElbowPlot(VascularObj, ndims = 50)
```


```{r}
### Determining the number of significant PCs
pct <- VascularObj@reductions$pca@stdev / sum(VascularObj@reductions$pca@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
vasc_sig_pcs <- 1:pcs
vasc_sig_pcs
```


```{r}
VascularObj <- FindNeighbors(VascularObj, dims = c(1:30))
```

```{r}
VascularObj <- RunUMAP(VascularObj, dims = c(1:30))
```


```{r}
DimPlot(VascularObj, group.by = "orig.ident")
```

```{r}
saveRDS(VascularObj, "~/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/VascularObj_Unintegrated.RDS")
```




```{r}
library(harmony)
VascularObj$samples <- as.factor(VascularObj$orig.ident)
VascularObj$samples <- as.character(VascularObj$samples)

Vascular_IntegratedObj <- RunHarmony(object = VascularObj,
                                     group.by.vars= "samples",
                                     verbose = T, assay.use= "RNA",max.iter.harmony = 100)

### Determining the number of significant PCs
pct <- Vascular_IntegratedObj@reductions$harmony@stdev / sum(Vascular_IntegratedObj@reductions$harmony@stdev) * 100
cum <- cumsum(pct)
co1 <- which(cum > 90 & pct < 5)[1]
co2 <- sort(which((pct[1:length(pct) -1] - pct[2:length(pct)]) > 0.05), decreasing = T)[1] + 1
pcs = min(co1, co2)
harmony_vasc_sig_pcs <- 1:pcs
harmony_vasc_sig_pcs

Vascular_IntegratedObj <- RunUMAP(Vascular_IntegratedObj,
                                  dims = c(1:30),
                                  reduction = "harmony", verbose = T)
```

## Quality of the integration
```{r}
DimPlot(Vascular_IntegratedObj, group.by = "orig.ident") + NoLegend()
```


```{r}
Vascular_IntegratedObj <- FindNeighbors(Vascular_IntegratedObj,
                                        dims = c(1:30),
                                        reduction = "harmony", verbose = T)
```


```{r}
res <- seq(0.1, 1, 0.1)
Vascular_IntegratedObj <- FindClusters(Vascular_IntegratedObj,
                                       resolution = res)
```


```{r, fig.width=12}
library(clustree)
clustree(Vascular_IntegratedObj)
```



```{r, fig.width=20, fig.height=10}
DimPlot(Vascular_IntegratedObj, 
         group.by = grep(pattern = "^RNA_snn", 
                         x = colnames(Vascular_IntegratedObj@meta.data)), 
         label = T, ncol = 5) & NoLegend()
```


## Signature plots
```{r}
ECs_markers_ref <- xlsx::read.xlsx("/home/mhassan/Files/Vascular_markers_summary.xlsx", sheetIndex = 1)

ECs_markers_ref <- lapply(ECs_markers_ref, function(x){
  x <- x[!(is.na(x))]
})
```


## Assiging signatures
```{r}
for (cell_type in names(ECs_markers_ref)) {
  
  # Access gene list for the current cell type
  current_gene_list <- ECs_markers_ref[[cell_type]]
  
  # Run AddModuleScore function on the Seurat object for the current cell type's gene list
  Vascular_IntegratedObj <- AddModuleScore(object = Vascular_IntegratedObj,
                               features = list(current_gene_list),
                               name = str_c(cell_type, "signature", sep = "_"),
                               search = T)
}
```


```{r}
colnames(Vascular_IntegratedObj@meta.data) <- gsub(pattern = "_signature1$", 
                                                   replacement = "_signature", 
                                                   x=colnames(Vascular_IntegratedObj@meta.data))
```

```{r}
pal <- viridis::viridis(n = 10, option = "C", direction = -1)
```


## Plotting signatures
```{r, fig.width=14, fig.height=14}
signature_columns <- colnames(Vascular_IntegratedObj@meta.data[grep(pattern = "_signature$",
                                                                    x = colnames(Vascular_IntegratedObj@meta.data))])

FeaturePlot(
  Vascular_IntegratedObj,
  features = signature_columns,
  cols = pal,
  ncol = 3
)
```


```{r, fig.height = 8}
FeaturePlot(Vascular_IntegratedObj, 
            features = c("PECAM1", "CDH5", "VWF", "MECOM"))
```


```{r}
DotPlot(Vascular_IntegratedObj, 
            features = c("PECAM1", "CDH5", "VWF", "MECOM"), group.by = "RNA_snn_res.0.2")
```

```{r}
Idents(Vascular_IntegratedObj) <- Vascular_IntegratedObj$RNA_snn_res.0.2
```


```{r}
DotPlot(Vascular_IntegratedObj, features = c("MRC1", "CFD", "IL7R", "PLIN1"))
```


```{r, fig.height = 8}
FeaturePlot(Vascular_IntegratedObj, features = c("MRC1", "CFD", "IL7R", "PLIN1"))
```



```{r, eval=FALSE}
DefaultAssay(Vascular_IntegratedObj) <- "RNA"
Idents(Vascular_IntegratedObj) <- Vascular_IntegratedObj$RNA_snn_res.0.2

markers_vascular_res02 <- FindAllMarkers(Vascular_IntegratedObj,
                              only.pos = T)

writexl::write_xlsx(as.data.frame(markers_vascular_res02), "~/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/markers_vasc_WRST_onlyPos_res02.xlsx")
```




```{r}
new_names = c(
'0' = 'V',
'1' = 'C',
'2' = 'A',
'3' = 'P',
'4' = 'SM',
'5' = 'P',
'6' = 'S',
'7' = 'S',
'8' = 'S',
'9' = 'LECs',
'10' = 'C'
)

Idents(Vascular_IntegratedObj) <- Vascular_IntegratedObj$RNA_snn_res.0.2

Vascular_IntegratedObj$vasc_annotation <- recode(Idents(Vascular_IntegratedObj),
                                                  !!!new_names)
```




```{r}
DimPlot(Vascular_IntegratedObj, group.by = "vasc_annotation", label = T, repel = T)
```


```{r}
saveRDS(Vascular_IntegratedObj, "~/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/Vascular_IntegratedObj.RDS")
```


```{r}
Idents(Vascular_IntegratedObj) <- "vasc_annotation"
markers_vasc_aft_ann <- FindAllMarkers(Vascular_IntegratedObj, only.pos = T)
writexl::write_xlsx(markers_vasc_aft_ann, "~/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/Vascular_markers_aft_ann.xlsx")
```



```{r}
save.image("~/GenomeDK/endoCellAtlas/AT_pipeline/Analysis/Doublets/scds/scds_Env.Rdata")
```





